R Notebook

### 2025-02-26 Scripts/authoritarian_only_latex_2.Rmd
Updates prior to CPS

---
output:
    bookdown::pdf_document2:
    keep_tex: true
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
indent: true
abstract: "The risk of mass protest figures prominently in many models of election manipulation, as a deterrent to tampering with elections. Yet election manipulation is common outside liberal democracies, while protest is rare (and successful protest rarer still). Using data on election fraud and duration data for protest events from 647 elections from 1990 to 2012, this paper argues that the risk to incumbents has been overstated in prior work. It argues that incumbents who solve the principal-agent dilemmas associated with election fraud are likely to be highly resilient to protest. There are two main findings. First, a riskier pre-election environment for incumbents is associated with more fraud, not less. Second, when protests do occur, greater election-day fraud is associated with a faster demise for protest movements. This pessimistic finding implies that election fraud is less risky than much prior work assumes, and helps explain the long-term stability of electoral authoritarianism as a regime type."
header-includes: 
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
  - \usepackage{multirow}
  - \usepackage{lscape}
  - \usepackage{pdflscape}

@@

Despite the significant challenge, the ruling party was able to defuse the protest movement over a period of months. Through a combination of repression and concessions [@golosov2012], the ruling party was able to reassert itself and continue a long process of sidelining the political opposition in Russia [@gelman_2015_troubled]. The post-2011 protest wave is one much-studied case of the phenomenon of electoral protest aimed at punishing electoral manipulation. Yet the Russian example highlights questions that get to the heart of a debate in the study of electoral manipulation and protest. When opposition movements protest manipulated election results, are they doing so because the incumbent\textemdash by relying on manufactured votes\textemdash has signaled its own weakness? This view is adopted by many researchers, especially in the formal modelling tradition. Or, does election manipulation indicate sources of strength for incumbents (independent of genuine popularity), as others have held?

The stakes of this debate are significant. The former view is an optimistic one; incumbents avoid fraud when it might be exposed and reveal their weakness to strategic protesters\textemdash an auspicious prediction for democratization and ultimately the under-pinning of 'self-enforcing' democracy [@fearon_self-enforcing_2011]. If the latter view holds, electoral protest may mainly occur non-strategically, and would typically fail to deter election manipulation as a result. Mass protest would, in this case, be a fragile guardrail against autocratization. This debate cannot be advanced by focusing solely on protest occurrence. Instead, this paper contributes by drawing out implications from three strands of the literature on election manipulation and protest, which have (to my knowledge) not previously been explicated.

In short: there are two views on the nature of electoral protest which make the same prediction. If opposition groups strategically initiate protest when they believe the incumbent to be weak, they will be most likely to do so when the election results show a low vote-share for the incumbent combined with high levels of manipulation\textemdash a combination of factors which reveals the incumbent's very low genuine popularity. [NOTE: ADD CITES HERE] However, if protest movements are more organic and non-strategic, reflecting citizen grievances, we should expect protest initiation in the same circumstances. Here the combination of low incumbent vote-share and high manipulation indicates severe citizen grievances (low natural popularity) which are galvanized by the shared grievance of a 'stolen' election [@tucker2007]. Since both theories make the same prediction about protest initiation, such tests\textemdash which have been the dominant empirical measure in the literature\textemdash do not reveal whether protest is strategic and manipulation indicates weakness or if manipulation is a hallmark of strong incumbents [@simpser2013] and protest is non-strategic.

To unravel this question, I take this analysis one step further. In strategic-protest models, the opposition protests because it believes the probability of a pro-government outcome of the protest is low. I assume that the duration of protest movements, when they begin, is an indicator of this underlying likelihood of a pro-government outcome. The more quickly a government can put down protests, the greater the underlying (unobserved) distribution of power favors the incumbent over the opposition. By contrast, the longer it takes the government to restore control, the less favorable the distribution of power.^[The analysis treats the small number of cases where the opposition succeeds in ousting the government in several ways. They are excluded from the main analysis, but incorporated in multiple robustness checks.]

I make use of data on the duration of post-election protests. Shorter-lived protest movements that end with the incumbent still in power indicate a government that is relatively strong compared to the protesting opposition, while longer-lived protest movements indicate a more even balance of power between the two. The reverse is also true: shorter-lived protest movements that end in opposition victory indicate a weaker government, while governments that hold on longer before being defeated are relatively stronger. Using data from 614 non-democratic elections across 97 countries, from 1989 to 2012, this paper finds that 1) protest initiation is most likely when manipulation is severe but incumbent vote-share is low, as both theories predict, and 2) that governments that engage in more manipulation are likely to put down protests more quickly than governments that produce little manipulation, especially when incumbents claim small vote-shares.

This pattern in the results is consistent with the argument that, rather than being a sign of weakness, the ability to generate significant electoral fraud is an indicator of durable regime strength. It is also consistent with the argument that much post-election protest in non-democracies is non-strategic; i.e., the actions of a weak opposition against a strong incumbent. Given these results, it is likely that the downside risk for incumbents that engage in election fraud in non-democracies has been overemphasized. This interpretation helps explain several empirical patterns at once: the relative paucity of electoral protests, the rarity of their success in overturning incumbent regimes, and governments' attempts to claim implausibly high margins of victory in conjunction with high levels of fraud. At a theoretical level, they also suggest that protest risk should be de-emphasized in research on the causes of election manipulation, in favor of alternative explanations which emphasize the role of front-line actors in delivering manipulation. Finally, they suggest a pessimistic outlook for democratization by election in authoritarian regimes.

@@

In classic models of election manipulation, manipulation is a short-term tool used instrumentally to secure victory in the election [@lehoucq2002]. Much of the work in this tradition has been elaborated through formal models. Though there are, of course, many variations, in several such models incumbents are characterized as 'weak' or 'strong' based on their true support in the electorate; weak incumbents may need to resort to fraud in order to secure re-election. As a result, the opposition perceives election fraud as a reliable signal of incumbent weakness, and will stage protests in order to push out the weak incumbent. For example, the model developed by Little, Tucker, and LaGatta [-@Little2015] predicts that "Just as stronger election results for the incumbent lead to less protest because they indicate that she is popular, reports of more fraud lead to more protest because they indicate that the incumbent is, for a given election result, less popular" (p. 1153). In equilibrium, this can lead weaker incumbents to step down rather than face protest due to massive fraud\textemdash a pro-democracy outcome. Similar results are developed by Chernykh and Svolik [-@chernykh2015third], Luo and Rozenas [-@Luo2020] and [-@Luo2017], and Magaloni [-@magaloni2010].

@@

Fundamentally, both set of models share the assumption that protest is a centralized, strategic decision by opposition groups, and that fraud is a tool employed by weak (i.e. unpopular) incumbents. As a result, they predict that the highest risk of protest should obtain when incumbents use massive fraud but still produce a low incumbent vote-share, indicating that the incumbent's true popularity is very low. This, in turn, indicates to opposition actors that protest is more likely to succeed in unseating the incumbent. These are normatively optimistic models upon which a sizable literature has been founded: fraud can be kept in check, and ultimately reduced, through the development of institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting]. These institutions improve election integrity, it is claimed, by increasing the likelihood that fraud will be observed and thus, incumbent weakness exposed.

@@

Most persuasively articulated by Simpser [-@simpser2013], this view holds that election manipulation demonstrates incumbent strength. As he writes, an incumbent that demonstrates the ability to engage in widespread manipulation "is also likely to have the resources, capacities, and inclinations to overcome or punish opponents, reward supporters, and circumvent the law" [@simpser2013, p. 6]. A strong incumbent (in the language used by the first set of models, i.e. one that is relatively popular) may nonetheless manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. By demonstrating otherwise difficult-to-observe capabilities, large-scale election manipulation can solve a coordination problem for the incumbent, securing the cooperation of elites [@simpser2013]. This can, in turn, fix expectations that the incumbent will continue to rule in the future, further enhancing its staying power [@hale_patronal_2014]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are associated with more election fraud.

For the set of theories that treat election manipulation as an indicator of sources of regime strength, manipulation can serve to deter protest under some circumstances. [NOTE ADD CITES HERE?] This leaves protest, when it does occur in the presence of large-scale manipulation, as something of a puzzle if protest behavior is largely strategic. For this reason, the manipulation-as-strength view comports more easily with another established view of the causes of electoral protest. Researchers in this tradition have argued that electoral protest occurs unrelated to strategic concerns about the relative strength of the incumbent and the opposition. In these models, electoral protest occurs because manipulation galvanizes citizens who hold diverse grievances, because economic conditions are poor, or simply because the bare fact of holding an election creates a moment of opportunity for opposition organization irrespective of the level of manipulation [@Pop-Eleches2015] [NOTE: ADD CITES]. The central feature of these models is that protest decisions are generally made without consideration for the likelihood that protests will overthrow the incumbent. As will be discussed in depth below, these different assumptions yield different predictions for when electoral protest is more likely to be initiated, and how long it will last against an incumbent that refuses to give up power.

This school of thought holds that decisions to engage in electoral protest are non-centralized, driven by citizens' grievances rather than by a strategic, organized opposition [@Brancati2016]. In this view, election manipulation can act as a shared grievance among individuals who prefer to see the ruling party lose\textemdash especially when it appears that fraud may have been decisive [@kuntz_more_2009]. Belief in a stolen election\textemdash one the incumbent would have lost in the absence of manipulation\textemdash is a particularly motivating kind of grievance, because it acts to focus all the diverse personal grievances individuals may have against the regime into one moment of opportunity for action [@tucker2007], possibly initiating a protest cascade [@Kuran1991]. These models thus tend to predict that protests are most likely when margins of victory are low and election manipulation high, suggesting both a high level of grievance in society and a "stolen" election. This framework predicts protest is most likely when incumbent vote-share is low and manipulation high (i.e. elevated grievances and a high likelihood of a stolen election).

@@

```{=latex}
% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\begin{tabular}{@{}lllllll@{}}
\toprule
                     &              &                           & \multicolumn{2}{l}{Strategic protest; manipulation = weakness} & \multicolumn{2}{l}{Non-strategic protest; manipulation = strength} \\ \midrule
Incumbent vote-share & Manipulation & True popularity inference & Initiation              & Prob. of pro-gov outcome             & Initiation                & Prob. of pro-gov. outcome              \\
Low                  & Low          & Low                       & Less likely             & More likely                          & Less likely               & Less likely                            \\
Low                  & High         & Very low                  & More likely             & Less likely                          & More likely               & More likely                            \\
Citations            &              &                           &                         &                                      &                           &                                        \\ \bottomrule
\end{tabular}
\caption{}
\label{tab:predictions}
\end{table}
```

Table 1 summarizes these predictions. As can be seen in the table, if we focus only on protest occurrence as a dependent variable, we cannot distinguish between two possible data-generating processes: one in which electoral protest is strategic and manipulation signals weakness, and one in which electoral protest is largely un-centralized but manipulation indicates regime strengths. Which of these data-generating processes predominates is of significant theoretical importance of understanding authoritarianism and democratization, since the latter view suggests much dimmer prospects for democratization than the former.

@@

Integrating Simpser's argument with recent research on the principal-agent dynamics inherent to election manipulation provides insights into this question. Where Simpser argues that manipulation should be blatant and excessive (that is, producing a large margin of victory) to send signals of strength, the principal-agent framework for understanding manipulation suggests otherwise. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver]. Factors that minimize principal-agent problems have been shown to facilitate greater election manipulation. These include the incumbent's local popularity [@rundlett2016deliver], the incumbent's ability to distribute patronage resources [@Harvey2019;@hale_patronal_2014], and low risk of punishment for pro-regime lawbreaking [@Harvey2022]. As discussed below, these are also independently useful for suppressing post-election protest. Consequently, if an incumbent is able to manufacture a large numbers of votes, it is also likely to have the resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country that can serve as independent sources of regime durability, even if the incumbent's nationwide margin of victory is low.

@@

Prior models of electoral protest by strategic opposition actors treat election manipulation and genuine votes as substitute goods. For any given vote-share, an increase in manufactured votes reduces the number of genuine votes. For classic models, which consider manufactured votes to be merely illusions of strength, the higher the number of manufactured votes for any given vote total, the weaker the incumbent. This means that strategic opposition actors are more likely to initiate protest when they perceive the incumbent to be weaker, and expect to be more likely to succeed in a challenge. A simplified version of this argument is presented in Panel B of Figure \@ref(fig:theoryfig). In the figure, dashed objects indicate variables that actors must estimate, while solid items are observed.

As the figure shows, in the classic model, the incumbent's (unknown) true popularity is positively related to its vote-share in the election, and negatively related to its use of fraud. This dynamic is the fulcrum upon which these models hinge: less popular incumbents (who are expected to be less likely to defeat an opposition protest) must manipulate more in order to win elections. Popular incumbents need manipulate less. Opposition actors gauge the probability of an pro-government protest outcome by estimating the incumbent's true popularity based on the degree of manipulation and the incumbent's vote-share in the election. The level of manipulation is, in effect, a mediator for the effect of vote-share. A high vote-share with low fraud indicates a more popular incumbent, likely deterring protest. A low vote-share with a high degree of fraud indicates a very weak incumbent, making strategic protest more likely. In other words, manipulated votes are empty in the classic model; subtracting them from the incumbent's vote-total gives opposition actors their estimate of the incumbent's ability to win a post-election standoff.

Thus the predicted relationship between manipulation and protest likelihood in the classic model is positive: more manipulation makes protest more likely. However, as increasing vote-share increases the number of genuine votes for the incumbent at any fixed level of fraud, the effect of fraud should be conditional on the claimed vote-share [NOTE: ADD CITES]. If the incumbent's total vote-share is 70 percent, for example, of which the opposition estimates 15 percent to be manufactured, the incumbent's genuine popularity is estimated at a respectable 55 percent. By contrast, if the incumbent posts a 50 percent margin of victory with the same 15 percent estimated to be manufactured, the opposition may perceive that the incumbent is weak in its genuine support (35 percent). This produces a first, baseline hypothesis, derived from the classic model.

It is important to note that the same hypothesis is also generated by non-centralized, non-strategic models of election protest. The mechanism, however, is different. In non-strategic models, protest emerges organically among disaffected citizens. Here, the combination of low incumbent vote-share and high manipulation is indicative of high grievances and a 'stolen' election. These are conditions that should be most likely to produce protest [@tucker2007]. By contrast, as vote-share increases for a set level of fraud, the incumbent's genuine popularity is higher and thus grievances are lower (reducing the likelihood of protest).

```{r theoryfig, echo=FALSE, fig.cap="A classic theory of election manipulation and protest", out.width="90%", fig.align='center'}
knitr::include_graphics("theory_comp.drawio.png")

@@

The classic model holds that manipulated votes correlate negatively with incumbents' post-election strength, implying that when incumbents rely heavily on election manipulation they will struggle to put down post-election protests. This effect should likewise be conditional on incumbent vote-share, since those incumbents who produce small margins of victory while relying heavily on manipulation will have limited ability to push back successfully against protest movements. Panel B in Figure [NOTE: FIX REF] illustrates this; a higher likelihood of pro-government protest outcome reduces the likelihood that the opposition will initiate protest. Consequently, classic models predict that more election manipulation should be associated with longer-lived protest movements against surviving incumbents.^[In the appendix, I check the separate prediction that manipulation should be positively associated with incumbents being ousted by protest movements.]

On the other hand, if election manipulation is an indicator of underlying capacities the incumbent may use to defeat protest movements, more severe electoral manipulation will be associated with shorter-lived protest waves. Panel A of Figure \@ref(fig:theoryfig) illustrates how the present argument differs from the classic model. Manufactured votes are not negatively associated with genuine popularity, but positively associated with impunity for agents and control over patronage resources. They are also positively associated with incumbent popularity [@rundlett2016deliver]. Abundant resources and extensive impunity reduce the principal-agent problems that frustrate election manipulation efforts. These same resources, institutions, and loyal agents make protest outcomes less likely to be favorable for the opposition. Note that, in this model, protest initiation is not tied to expected protest outcome; protest is driven by non-strategic concerns related to the incumbent's popularity (i.e. citizens' diverse grievances) and whether manipulation is perceived as having been decisive (i.e. a stolen election). Incumbents with the resources and institutions needed to produce widespread election manipulation can draw on those same capacities to buy off, repress, or otherwise quash post-election protest. This yields Hypothesis 2.

@@

p.hist.events <- ggplot(nelda2.sub, aes(x=n.events.post)) + geom_histogram(bins = 90) + xlab("Number of protest events") + theme_bw()

p.hist.duration <- ggplot(nelda2.sub, aes(x=d.events.num)) + geom_histogram(bins=100) + xlab("Protest wave duration") + theme_bw()

>>>>

---
output:
    bookdown::pdf_document2:
    keep_tex: true
linkcolor: blue
bibliography: C:/Users/colej/Documents/Research projects/bib_ch.bib
#biblio-style: "apalike"
fontsize: 12pt
link-citations: yes
toc: false
csl: C:/Users/colej/Documents/Research projects/ji_manipulation_protest/american-journal-of-political-science.csl
indent: true
abstract: "Two models of the relationship between election manipulation and protest can be found in the literature. A classic approach holds that election fraud reveals regime weakness, which can lead to strategic opposition protest. To avoid costly protest, incumbents may hold cleaner elections. A more recent approach contends that election manipulation can be a sign of underlying regime strengths; under this framework protest is less likely to deter manipulation. Prior empirical tests do not cleanly distinguish between these models. Using event-day data on electoral protest in 445 authoritarian cases from 1990-2012, this paper shows that 1) fraud severity is rarely associated with protest initiation, and 2) governments which manipulate more also defeat protests more quickly. This goes against the classic, deterrent model, and helps explain why electoral authoritarianism is a sticky equilibrium amid ongoing democratic backsliding. It suggests researchers should look beyond protest risk in models of electoral manipulation."
header-includes: 
  - \usepackage{booktabs}
  - \usepackage{setspace}
  - \usepackage{float}
  - \usepackage{caption}
  - \captionsetup[table]{labelformat=empty}
  - \usepackage{multirow}
  - \usepackage{lscape}
  - \usepackage{pdflscape}

@@

Despite the significant challenge, the ruling party was able to defuse the protest movement over a period of months. Through a combination of repression and concessions [@golosov2012], the ruling party was able to reassert itself and continue a long process of sidelining the political opposition in Russia [@gelman_2015_troubled]. The post-2011 protest wave is one much-studied case of the phenomenon of electoral protest aimed at punishing electoral manipulation. Yet the Russian example highlights questions that get to the heart of a debate in the study of electoral manipulation and protest. When opposition movements protest manipulated election results, are they doing so because the incumbent\textemdash by relying on manufactured votes\textemdash has signaled its own weakness? This view is adopted by many researchers, especially in the formal modelling tradition. Or, does election manipulation indicate sources of strength for incumbents (independent of genuine popularity), and does protest occur for non-strategic reasons?

The stakes of this debate are significant. The former view is an optimistic one; in it, incumbents may restrain fraud when it might be exposed and reveal their weakness to strategic protesters\textemdash an auspicious prediction for democratization and ultimately the under-pinning of 'self-enforcing' democracy [@fearon_self-enforcing_2011]. If the latter view holds, non-strategic electoral protest may fail to deter election manipulation. The risk of mass protest would, in this case, be a fragile guardrail against autocratization. This debate cannot be advanced by focusing solely on protest occurrence. Instead, this paper contributes by drawing out implications from three strands of the literature on election manipulation and protest, which have (to my knowledge) not previously been explicated.

In short: there are two views on the nature of electoral protest which make the same prediction. If opposition groups strategically initiate protest when they believe the incumbent to be weak, they will be most likely to do so when the election results show a low vote-share for the incumbent combined with high levels of manipulation\textemdash a combination of factors which reveals the incumbent's very low genuine popularity [@Luo2017;@Little2012;@Schedler2002;@Beaulieu2014a]. However, if protest movements are more organic and non-strategic, reflecting citizen grievances, we should expect protest initiation in the same circumstances. Here the combination of low incumbent vote-share and high manipulation indicates severe citizen grievances (low natural popularity) which are galvanized by the shared grievance of a 'stolen' election [@tucker2007]. Since both theories make the same prediction about protest initiation, such tests\textemdash which have been the dominant empirical measure in the literature\textemdash do not reveal whether protest is strategic and manipulation indicates weakness or if manipulation is a hallmark of strong incumbents [@simpser2013] and protest is non-strategic.

To unravel this question, I take this analysis one step further. In strategic-protest models, the opposition protests because it believes the probability of a pro-government outcome of the protest is low. I assume that the duration of protest movements, when they occur, is an indicator of this latent likelihood of a pro-government outcome. The more quickly a government puts down protests, the greater the underlying (unobserved) distribution of power favors the incumbent over the opposition. By contrast, the longer it takes the government to restore control, the less favorable the distribution of power.^[The analysis treats the small number of cases where the opposition succeeds in ousting the government in several ways. They are excluded from the main analysis, but incorporated in a robustness check.]

Using data from 445 non-democratic elections across 97 countries, from 1989 to 2012, this paper finds that 1) protest initiation is most likely when manipulation is severe but incumbent vote-share is low, as both theories predict, and 2) that governments that engage in more manipulation are likely to put down protests more quickly than governments that produce little manipulation, especially when incumbents claim small vote-shares.

This pattern in the results is consistent with the argument that, rather than being a sign of weakness, the ability to generate significant electoral fraud is an indicator of durable regime strength. It is also consistent with the argument that much post-election protest in non-democracies is non-strategic; i.e., the actions of a weak opposition against a strong incumbent. Given these results, it is likely that\textemdash while incumbents do not seek to provoke protest\textemdash the downside risk for incumbents that engage in election fraud in non-democracies has been overemphasized. This interpretation helps explain several empirical patterns at once: the relative paucity of electoral protests [@Brancati2016], the rarity of their success in overturning incumbent regimes, and governments' attempts to claim implausibly high margins of victory in conjunction with high levels of fraud. At a theoretical level, they also suggest that protest risk should be de-emphasized in research on the causes of election manipulation, in favor of alternative explanations which emphasize the role of front-line actors in delivering manipulation. Finally, they suggest a pessimistic outlook for democratization by election in authoritarian regimes.

@@

In classic models of election manipulation, manipulation is a short-term tool used instrumentally to secure victory in the election [@lehoucq2002]. Much of the work in this tradition has been elaborated through formal models. Though there are, of course, many variations, in several such models incumbents are characterized as 'weak' or 'strong' based on their true support in the electorate. This true level of support can be obscured prior to elections due to pervasive preference falsification [@Kuran1991] and the low level of political information common in authoritarian regimes [@guriev2020]. However, low popularity (i.e. weakness, in these models) can be revealed when incumbents need to resort to fraud in order to secure re-election. As a result, the opposition perceives election fraud as a reliable signal of incumbent weakness, and will stage protests in order to push out the weak incumbent. For example, the model developed by Little, Tucker, and LaGatta [-@Little2015] predicts that "Just as stronger election results for the incumbent lead to less protest because they indicate that she is popular, reports of more fraud lead to more protest because they indicate that the incumbent is, for a given election result, less popular" (p. 1153). In equilibrium, this can lead weaker incumbents to step down rather than face protest due to massive fraud\textemdash a pro-democracy outcome. Similar results are developed by Chernykh and Svolik [-@chernykh2015third], Luo and Rozenas [-@Luo2020] and [-@Luo2017], and Magaloni [-@magaloni2010].

@@

Fundamentally, both set of models share the assumption that protest is a centralized, strategic decision by opposition groups, and that fraud is a tool employed by weak (i.e. unpopular) incumbents. As a result, they predict that the highest risk of protest should obtain when incumbents use massive fraud but still produce a low incumbent vote-share, indicating that the incumbent's true popularity is very low. This, in turn, indicates to opposition actors that protest is more likely to succeed in unseating the incumbent. Incumbents in these models typically face a tradeoff between the short-term benefits of electoral manipulation for winning the election, and any weakness that relying on fraud reveals to the opposition [@Rozenas2016b]. As Beaulieu [-@Beaulieu2014a, p. 48] writes in this tradition, "As the opposition grows stronger, the amount of [manipulation] he will tolerate shrinks, and the incumbent who wishes to avoid protest will need to reduce the amount of manipulation she commits." These are normatively optimistic models upon which a sizable literature has been founded: fraud can be kept in check, and ultimately reduced, through the development of institutions that reveal information about election integrity: election monitoring [@Hyde2014;@daxecker2012], more independent courts [@chernykh2015third], media activity [@naiFourthEstate], and civil society coordination [@birch2017getting]. These institutions improve election integrity, it is claimed, by increasing the likelihood that fraud will be observed and thus, incumbent weakness exposed.

@@

Most persuasively articulated by Simpser [-@simpser2013], this view holds that election manipulation demonstrates incumbent strength. As he writes, an incumbent that demonstrates the ability to engage in widespread manipulation "is also likely to have the resources, capacities, and inclinations to overcome or punish opponents, reward supporters, and circumvent the law" [@simpser2013, p. 6]. A relatively popular incumbent ("strong" in the language used by the first set of models) may nonetheless manipulate heavily in order to produce a very large majority [@magaloni2006], to induce regime insiders to remain cooperative [@gehlbach2015electoral], or because of a bandwagon effect among front-line agents [@rundlett2016deliver]. By demonstrating otherwise difficult-to-observe capabilities, large-scale election manipulation can solve a coordination problem for the incumbent, securing the cooperation of elites [@simpser2013]. This can, in turn, fix expectations that the incumbent will continue to rule in the future, further enhancing its staying power [@hale_patronal_2014]. This argument is in keeping with Seeberg's [-@Seeberg2019] finding that higher-capacity states are associated with more election fraud.

If election manipulation can be an indicator of regime strength, it may serve to deter protest under some circumstances, rather than the other way around [@simpser2013]. This leaves protest, when it does occur in the presence of large-scale manipulation, as something of a puzzle if protest behavior is largely strategic. For this reason, the manipulation-as-strength view comports more easily with another established view of the causes of electoral protest. Researchers in this tradition have argued that electoral protest occurs largely unrelated to strategic concerns about the relative strength of the incumbent and the opposition. In these models, electoral protest occurs because manipulation galvanizes citizens who hold diverse grievances, because economic conditions are poor, or simply because the bare fact of holding an election creates a moment of opportunity for opposition organization irrespective of the level of manipulation [@Pop-Eleches2015]. The central feature of these models is that protest decisions are generally made without consideration for the likelihood that protests will overthrow the incumbent.^[To be clear, this is the narrow sense in which 'strategic' and 'non-strategic' is used here and in protest-deterrent models. Protest behavior may still be rational in a broader sense without being narrowly tied to the expected likelihood of ousting the government. For instance, protesters may believe it is unlikely they will be individually punished for participating in protests, making it individually rational for them to voice their grievances.]

This school of thought holds that decisions to engage in electoral protest are driven in large part by citizens' grievances rather than by a strategic opposition [@tucker2007]. In this view, election manipulation can act as a shared grievance among individuals who prefer to see the ruling party lose\textemdash especially when it appears that fraud may have been decisive [@kuntz_more_2009]. Belief in a stolen election\textemdash one the incumbent would have lost in the absence of manipulation\textemdash is a particularly motivating kind of grievance, because it acts to focus all the diverse personal grievances individuals may have against the regime into one moment of opportunity for action [@tucker2007], possibly initiating a protest cascade [@Kuran1991]. These models thus tend to predict that protests are most likely when margins of victory are low and election manipulation high, suggesting both a high level of grievance in society and a "stolen" election. This framework predicts protest is most likely when incumbent vote-share is low and manipulation high (i.e. elevated grievances and a high likelihood of a stolen election).

Rational protest entrepreneurs might seek to initiate protest to take advantage of this popular discontent for non-strategic reasons, in order to press for improved competitiveness in the *next* election or to bargain for policy concessions [@chernykh2014;@robertson_protest_2010]. They may stage protests in order to identify new participants, forge new social ties [@barberaModelProtestsRevolution2019] and test new tactics. They may seek to communicate information about the government's unpopularity, to set the stage for larger, future protests [@lohmannDynamicsInformationalCascades1994]. Protests may be staged as part of internal factional rivalries within the opposition [@Buckles2017], or to drive media coverage of issues important to the opposition [@huangMediaProtestDiffusion2019;@jenningsStreetDemonstrationsMedia2019]. Or, a vanguard of protesters may take to the streets simply because their distaste for the government outweighs any risk of sanction [@lohmannDynamicsInformationalCascades1994].  These and other motivations for initiating protest are theoretically significant because their rationality (for protesters) does not hinge on the likelihood of overturning the government, as classic models assume. As a result, it is possible that we may observe the initiation of anti-government electoral protest even when the likelihood of such protest overturning the election (or the government itself) is very low.

@@

Taking this all together, it becomes clear that if researchers focus only on protest occurrence as a dependent variable, they cannot distinguish between two possible data-generating processes: one in which manipulation signals weakness and electoral protest is a strategic response, and one in which electoral manipulation indicates regime strengths and protest occurs for a variety of reasons with limited connection to the odds of removing the government. In the former, protest is predicted to occur most often when manipulation is high and the incumbent's vote-share is low, since this provides a strong signal that the incumbent's true popularity is low. In the latter, protest is most likely in the same condition because the incumbent's low popularity indicates widespread grievances, which the high level of manipulation serves to coordinate despite the incumbent's organizational strengths. Which of these data-generating processes predominates is of significant theoretical importance of understanding authoritarianism and democratization, since the latter view suggests much dimmer prospects for democratization than the former.

@@

Prior models of electoral protest by strategic opposition actors, discussed above, treat election manipulation and genuine votes as substitute goods. For any given vote-share, an increase in manufactured votes reduces the number of genuine votes. For classic models, which consider manufactured votes to be merely illusions of strength, the higher the number of manufactured votes for any given vote total, the weaker the incumbent. This means that strategic opposition actors are more likely to initiate protest when they perceive the incumbent to be weaker, and expect to be more likely to succeed in a challenge.

In the classic model, the incumbent's (unknown) true popularity is positively related to its vote-share in the election, and negatively related to its use of fraud. This dynamic is the fulcrum upon which these models hinge: less popular incumbents (who are expected to be less likely to defeat an opposition protest) must manipulate more in order to win elections. Popular incumbents need manipulate less. Opposition actors gauge the probability of an pro-government protest outcome by estimating the incumbent's true popularity based on the degree of manipulation and the incumbent's vote-share in the election. A high vote-share with low fraud indicates a more popular incumbent, likely deterring protest. A low vote-share with a high degree of fraud indicates a very weak incumbent, making strategic protest more likely. In other words, manipulated votes are empty in the classic model; subtracting them from the incumbent's vote-total gives opposition actors their estimate of the incumbent's ability to win a post-election standoff.

Thus the predicted relationship between manipulation and protest likelihood in the classic model is positive: more manipulation makes protest more likely. However, as increasing vote-share increases the number of genuine votes for the incumbent at any fixed level of fraud, the effect of fraud should be conditional on the claimed vote-share. As Rozenas [@Rozenas2016b] notes, if an incumbent wins an election with low manipulation, it generates a strong signal of its popularity; if it wins a heavily manipulated election, this signal is weaker. As an illustration, if the incumbent's total vote-share is 70 percent, for example, of which the opposition estimates 15 percent to be manufactured, the incumbent's genuine popularity is estimated at a respectable 55 percent. By contrast, if the incumbent posts a 50 percent margin of victory with the same 15 percent estimated to be manufactured, the opposition may perceive that the incumbent is weak in its genuine support (35 percent). This produces a first, baseline hypothesis, derived from the classic model.

It is important to note that the same hypothesis is also generated by non-centralized, non-strategic models of election protest. The mechanism, however, is different. Here, the combination of low incumbent vote-share and high manipulation is indicative of widespread grievances coordinated a 'stolen' election. These are conditions that should be most likely to produce protest [@tucker2007]. By contrast, as vote-share increases for a set level of fraud, the incumbent's genuine popularity is higher and thus grievances are lower (reducing the likelihood of protest).

Integrating Simpser's argument with recent research on the principal-agent dynamics inherent to election manipulation provides different insights. Where Simpser argues that manipulation should be blatant and excessive (that is, producing a large margin of victory) to send signals of strength, the principal-agent framework for understanding manipulation suggests otherwise. This research trend emphasizes the degree to which political principals require the cooperation of large numbers of front-line agents in order to generate extensive election fraud, agents who may shirk if their perceived interests are not aligned with the incumbent [@rundlett2016deliver]. Factors that minimize principal-agent problems have been shown to facilitate greater election manipulation. These include the incumbent's local popularity [@rundlett2016deliver], the incumbent's ability to distribute patronage resources [@Harvey2019;@hale_patronal_2014], and low risk of punishment for pro-regime lawbreaking [@Harvey2022]. As discussed below, these are also independently useful for suppressing post-election protest. Consequently, if an incumbent is able to manufacture a large numbers of votes, it is also likely to have the resource capacity, control over bureaucratic institutions, and perhaps deep wells of popularity in parts of the country that can serve as independent sources of regime durability, even if the incumbent's nationwide margin of victory is low.

@@

If election manipulation is an indicator of underlying capacities the incumbent may use to defeat protest movements, more severe electoral manipulation will be associated with shorter-lived protest waves. Manufactured votes are not negatively associated with genuine popularity, but positively associated with impunity for agents and control over patronage resources. They are also positively associated with incumbent popularity [@rundlett2016deliver]. Abundant resources and extensive impunity reduce the principal-agent problems that frustrate election manipulation efforts. These same resources, institutions, and loyal agents make protest outcomes less likely to be favorable for the opposition. Note that, in this model, protest initiation is not tied to expected protest outcome; protest is driven by non-strategic concerns related to the incumbent's popularity (i.e. citizens' diverse grievances) and whether manipulation is perceived as having been decisive (i.e. a stolen election). Incumbents with the resources and institutions needed to produce widespread election manipulation can draw on those same capacities to buy off, repress, or otherwise quash post-election protest. This yields Hypothesis 2.

```{r theoryfig, echo=FALSE, fig.cap="A classic theory of election manipulation and protest", out.width="90%", fig.align='center', include=FALSE}
knitr::include_graphics("theory_comp.drawio.png")

@@

The classic model holds that manipulated votes correlate negatively with incumbents' post-election strength, implying that when incumbents rely heavily on election manipulation they will struggle to put down post-election protests. This effect should likewise be conditional on incumbent vote-share, since those incumbents who produce small margins of victory while relying heavily on manipulation will have limited ability to push back successfully against protest movements. A higher likelihood of pro-government protest outcome reduces the likelihood that the opposition will initiate protest. Consequently, classic models predict that more election manipulation should be associated with longer-lived protest movements against surviving incumbents.^[In the appendix, I check the separate prediction that manipulation should be positively associated with incumbents being ousted by protest movements.]

@@

p.hist.events <- ggplot(nelda2.sub %>% filter(inc.lose == "0"), aes(x=n.events.post)) + geom_histogram(bins = 90) + xlab("Number of protest events") + theme_bw()

p.hist.duration <- ggplot(nelda2.sub %>% filter(inc.lose == "0"), aes(x=d.events.num)) + geom_histogram(bins=100) + xlab("Protest wave duration") + theme_bw()
